PyUltraLight: A Pseudo-Spectral Solver for Ultralight Dark Matter Dynamics

### 2018-05-27 draft.tex
small changes from Richard

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}

@@

\newcommand{\apj}{Astrophysical J.}
\newcommand{\prd}{Phys. Rev. D}
\newcommand{\mnras}{Mon. Notices Royal Astron. Soc}
\newcommand{\PyUltraLight}{\textsc{PyUltraLight }}

\abstract{\PyUltraLight\ simulates the dynamics of ultralight dark matter in a static, non-expanding background.  \PyUltraLight\  can describe the evolution of several interacting ultralight dark matter halos  or one or more halos orbiting a central, fixed Newtonian potential, the latter scenario corresponding to dwarf galaxies orbiting a massive central galaxy. We verify  \PyUltraLight\  by showing that it reproduces qualitative dynamical features of previously published simulations and demonstrate that it has  excellent energy-conservation properties.   \PyUltraLight\  is implemented in a Python-based Jupyter notebook, solving the Schr{\"o}dinger-Poisson equation  governing  ultralight scalar field dark matter dynamics in the non-relativistic regime using a symmetrised split-step pseudo-spectral  algorithm. The notebook interface makes it simple to specify simulation parameters and  visualise the resulting output. \PyUltraLight\  runs on standard desktop hardware and is available on GitHub.  }

@@

The realisation that there may be more to the universe than meets the eye is one of the most profound developments in 20$^{\textrm{th}}$ Century astronomy and astrophysics. However,  while there are now multiple lines of evidence that dark matter outweighs baryonic matter by a ratio of approximately 5:1 [CITE], we have few clues regarding the physical nature of dark matter.  Much theoretical and experimental effort has focussed on  WIMP models, motivated by their consistency with supersymmetric extensions to the Standard Model and their relatively simple dynamics but advanced direct-detection experiments are putting increasingly tight constraints on the WIMP parameter space \cite{Tan2016, Akerib2017}. Moreover, $\Lambda$CDM cosmology with simple, pressureless, noninteracting dark matter (a class which includes simple WIMP models) is potentially at odds with observations at small astrophysical scales \cite{Bull2016}.

[TO HERE]The potential shortcoming of simple cold dark matter scenarios are increasingly motivating the investigations of more novel dark matter scenarios. In particular, ultralight dark matter (ULDM) is an increasingly well-studied possibility, also known as fuzzy dark matter (FDM), or BEC dark matter. ULDM models are well motivated by fundamental theories possessing approximate shift symmetries, and offer natural resolutions to the small-scale problems of $\Lambda$CDM without the need to invoke baryonic physics. This is essentially because below the de Broglie wavelength of the ULDM particle, the Heisenberg uncertainty principle suppresses gravitational collapse. Since the mass of the ULDM particle is postulated to be of order $10^-22$ eV, this amounts to suppression of gravitational collapse on kilo parsec scales \cite{Hu2000}. A recent review of ULDM/FDM is given in \cite{Hui2016}.

The computational modelling of ULDM systems in the non-relativistic regime can be distilled into numerically solving the dynamical Schr{\"o}dinger-Poisson (SP) system of coupled differential equations for given initial conditions. Approaches to this problem have been varied, and have given rise to modifications of existing cosmological simulation codes as well as the development of new codes specifically for the modelling of ULDM systems, though public releases of the latter are  lacking. One approach has been to use the Madelung fluid formulation of the SP system \cite{Madelung1926}, in which there arises a quantum pressure term which can be treated numerically in a variety of ways. In \cite{Zhang2018}, the cosmological code \textsc{gadget} \cite{Springel2005} has been modified in order to treat this quantum pressure using an effective particle-particle interaction. This modified code, \textsc{axion-gadget}, has been made publicly available \cite{axion-gadget}. In \cite{Nori2018}, a non-public extension of \textsc{gadget}, \textsc{p-gadget3}, has been modified to treat the quantum pressure term using smoothed-particle hydrodynamics (SPH) routines. The SPH approach has also been used in \cite{Mocz2015}, while a particle-mesh approach was implemented in \cite{Veltmaat2016} to treat the quantum pressure term. Other cosmological simulation codes which have been recruited to solve ULDM dynamics (though not necessarily using the Madelung formulation) are N\textsc{yx} \cite{Almgren2013}, G\textsc{alacticus} \cite{Benson2012}, \textsc{arepo} \cite{Springel2010}, and \textsc{gamer} \cite{Schive2010, gamer}. N\textsc{yx} was modified in \cite{Schwabe2016} to study merging ULDM solitonic cores, G\textsc{alacticus} was modified in \cite{Du2017} to study the effects of tidal stripping and dynamical friction on ULDM halos, \textsc{arepo} was modified in \cite{Mocz2017} to study the core-mass relationship and turbulence characteristics of ULDM halos, and \textsc{gamer} was modified in \cite{Schive2014_b} to perform a detailed study of structure formation in ULDM cosmologies.

>>>>

\usepackage[T1]{fontenc} % if needed
\usepackage{eucal}
\usepackage{xspace}
\usepackage{color}

@@

\newcommand{\apj}{Astrophysical J.}
\newcommand{\prd}{Phys. Rev. D}
\newcommand{\mnras}{Mon. Notices Royal Astron. Soc}
\newcommand{\PyUltraLight}{\textsc{PyUltraLight}\xspace}
\newcommand{\re}[1]{\textcolor{blue}{[{\bf RE}: #1]}}

\abstract{\PyUltraLight  simulates the dynamics of ultralight dark matter in a static, non-expanding background.  \PyUltraLight can describe the evolution of several interacting ultralight dark matter halos  or one or more halos orbiting a central, fixed Newtonian potential, the latter scenario corresponding to dwarf galaxies orbiting a massive central galaxy. We verify  \PyUltraLight  by showing that it reproduces qualitative dynamical features of previously published simulations and demonstrate that it has  excellent energy-conservation properties.   \PyUltraLight is implemented in a Python-based Jupyter notebook, solving the Schr{\"o}dinger-Poisson equation  governing  ultralight scalar field dark matter dynamics in the non-relativistic regime using a symmetrised split-step pseudo-spectral  algorithm. The notebook interface makes it simple to specify simulation parameters and  visualise the resulting output. \PyUltraLight  runs on standard desktop hardware and is available on GitHub.  }

@@

The realisation that there may be more to the universe than meets the eye is one of the most profound developments in 20$^{\textrm{th}}$ Century astronomy and astrophysics. However,  while there are now multiple lines of evidence that dark matter outweighs baryonic matter by a ratio of approximately 5:1 \re{Add a citation to this}, we have few clues regarding the physical nature of dark matter.  Much theoretical and experimental effort has focussed on  WIMP models, motivated by their consistency with supersymmetric extensions to the Standard Model and their relatively simple dynamics. However but advanced direct-detection experiments are putting increasingly tight constraints on the WIMP parameter space \cite{Tan2016, Akerib2017} and $\Lambda$CDM cosmology with simple, pressureless, noninteracting dark matter (a class  including simple WIMP scenarios) is potentially at odds with observations at small astrophysical scales \cite{Bull2016}.

The potential shortcoming of simple cold dark matter scenarios   motivate  investigations of more novel dark matter scenarios. In particular, ultralight dark matter (ULDM), also known as fuzzy dark matter (FDM), or BEC dark matter, is an increasingly well-studied possibility; for a recent review of the potential advantages and characteristic attributes of this scenario see Ref~\cite{Hui2016}.   ULDM models are well motivated by fundamental theories possessing approximate shift symmetries. Moreover, ULDM can naturally resolve the small-scale problems of $\Lambda$CDM as the Heisenberg uncertainty principle suppresses gravitational collapse on lengthscales shorter than the de Broglie wavelength of the ULDM particle. In this regime the mass of the ULDM particle becomes correlated with astrophysical observables; if it is on the order of $10^{-22}$~eV, structure is suppressed at kiloparsec scales and below \cite{Hu2000}.

\re{read to here}The computational modelling of ULDM systems in the non-relativistic regime can be distilled into numerically solving the dynamical Schr{\"o}dinger-Poisson (SP) system of coupled differential equations for given initial conditions. Approaches to this problem have been varied, and have given rise to modifications of existing cosmological simulation codes as well as the development of new codes specifically for the modelling of ULDM systems, though public releases of the latter are  lacking. One approach has been to use the Madelung fluid formulation of the SP system \cite{Madelung1926}, in which there arises a quantum pressure term which can be treated numerically in a variety of ways. In \cite{Zhang2018}, the cosmological code \textsc{gadget} \cite{Springel2005} has been modified in order to treat this quantum pressure using an effective particle-particle interaction. This modified code, \textsc{axion-gadget}, has been made publicly available \cite{axion-gadget}. In \cite{Nori2018}, a non-public extension of \textsc{gadget}, \textsc{p-gadget3}, has been modified to treat the quantum pressure term using smoothed-particle hydrodynamics (SPH) routines. The SPH approach has also been used in \cite{Mocz2015}, while a particle-mesh approach was implemented in \cite{Veltmaat2016} to treat the quantum pressure term. Other cosmological simulation codes which have been recruited to solve ULDM dynamics (though not necessarily using the Madelung formulation) are N\textsc{yx} \cite{Almgren2013}, G\textsc{alacticus} \cite{Benson2012}, \textsc{arepo} \cite{Springel2010}, and \textsc{gamer} \cite{Schive2010, gamer}. N\textsc{yx} was modified in \cite{Schwabe2016} to study merging ULDM solitonic cores, G\textsc{alacticus} was modified in \cite{Du2017} to study the effects of tidal stripping and dynamical friction on ULDM halos, \textsc{arepo} was modified in \cite{Mocz2017} to study the core-mass relationship and turbulence characteristics of ULDM halos, and \textsc{gamer} was modified in \cite{Schive2014_b} to perform a detailed study of structure formation in ULDM cosmologies.
